---
title: "Structural Zeros in Correlation Matrices: A Twelve-Parameterization Bake-Off"
author: "Sean Pinkney"
date: "2026-07-15"
categories: [stan, correlation, parameterizations]
bibliography: references.bib
format:
  html:
    code-fold: true
---

## 1. The Problem with Zeros in Correlation Matrices

Sometimes you know, before seeing any data, that certain correlations are exactly zero. A graphical model tells you two variables are conditionally unrelated, a confirmatory factor structure forbids a cross-loading, a physical system rules out a coupling. The prior you want is "LKJ [@lewandowski2009], restricted to the matrices that respect my zeros."

You cannot get there by sampling from the LKJ and zeroing out cells, as the result is generally not positive definite. The zeros have to be built into the parameterization itself, and how you build them in turns out to matter a great deal for the sampler. In this post I race twelve Stan [@carpenter2017] parameterizations of the same target posterior against each other. Since all twelve provably target the same distribution (and I verify this numerically), the comparison isolates exactly one thing: **sampling geometry**. Divergences, effective sample size, and cost per gradient evaluation.

The punchline: the two families of solutions make opposite trade-offs, and a hybrid, an exact-constraint construction with a canonical triangular basis, refuses to trade at all. **Zero divergences in every run, the best ESS per gradient in the field, and the fastest wall clock.**

## 2. Family 1: Forced Zeros

Family 1 is the constrained-Cholesky approach I introduced in @pinkney2024 and presented at StanCon [@pinkneytalk]. The original method is more general than zeros: it walks the Cholesky factor row by row and lets you pin any cell to a *known value* or restrict it to a *bounded range*. Writing $\Omega = L L'$ with unit-norm rows of $L$, the correlation in cell $(i,j)$ is $$
\Omega_{ij} = b_1 + L_{ij} L_{jj}, \qquad
b_1 = \sum_{k<j} L_{ik} L_{jk},
$$ which is affine in $L_{ij}$, so a known value — here zero — pins $L_{ij} = -b_1 / L_{jj}$ and consumes no parameter, while a bound on $\Omega_{ij}$ becomes a bound on $L_{ij}$. Conditioning the LKJ distribution on $\Omega_{ij} = 0$, a delta function in $\Omega$ coordinates, contributes a density correction of $-\log L_{jj}$ per zero. The `carpenter` variant below is Bob Carpenter's compact formulation of the zeros-only case, from a [recent Stan Discourse exchange](https://discourse.mc-stan.org/t/new-users-guide-intro-sem-and-user-defined-constrained-parameters-chapters/41430) about exactly this problem:

``` {.stan}
if (is_zero(i, j, zeros, zero_idx)) {
  real b = dot_product(L[j, 1:(j - 1)], L[i, 1:(j - 1)]);
  L[i, j] = -b / L[j, j];   // implies Omega[i, j] == 0
  jacobian += -log(L[j, j]);
  zero_idx += 1;
}
```

The forced-zero trick composes with essentially any Cholesky-style parameterization of the free cells, which is how one idea becomes eight models:

| label | free-cell parameterization |
|:---|:---|
| `carpenter` | direct Cholesky entries, bounded by the remaining stick |
| `pinkney` | correlation-scale value divided by $L_{jj}$, multiplicative stick [@pinkney2024; @pinkneytalk] |
| `ldl` / `ldl2` / `ldl3` | $LDL'$ factorization [@pinkney2024]; stick tracked via `log_diff_exp`, a `log1m` ratio, or the natural scale |
| `cvine` | C-vine partial correlations with the LKJ Beta factorization [@joe2006; @lewandowski2009] |
| `rowscale` | direct Cholesky entries, row-scaling stick |
| `schur` | canonical partial correlations, $L_{ij} = r \tanh(x)$ |

They all share one flaw. The forced value $-b_1/L_{jj}$ does not know about the row's remaining stick, $\sqrt{1 - \sum_{k<j} L_{ik}^2}$, and nothing stops it from overshooting. When it does, a downstream `sqrt` goes negative, the log density is `NaN`, and the sampler is suddenly fighting a hard boundary it cannot see coming. Every member of this family runs at roughly a **1% divergence rate** on the benchmarks below. That is a bias risk you manage with `adapt_delta` forever, on every model you ever build with these blocks.

Before moving on, a measurement lesson that the LDL trio taught me for free. The three variants define *identical* densities and differ only in floating-point arithmetic, yet their divergence totals across five patterns span 356 to 466. That spread is a direct read on the noise floor of single-fit divergence counts, and it is wider than most gaps between genuinely different parameterizations. **Beware of divergence league tables built from single runs.**

## 3. Family 2: Exact-Constraint Subspaces

The second idea eliminates the boundary instead of policing it. I learned it from Seth Axen, in a Turing.jl discussion on positive definite matrices with structural zeros [@axen2023]:

> For each column j, each column i for i \< j that is orthogonal to column j decreases the dimensionality of the hemisphere by 1. So we simply construct a point on a lower dimensional hemisphere and use the QR decomposition of the orthogonal columns to lift it to a higher dimensional hemisphere.

Row $i$ of a correlation Cholesky factor is a unit vector with a positive last coordinate. Because row $j$ is zero beyond column $j$, the constraint $\Omega_{ij} = 0$ says row $i$ is **orthogonal to row** $j$ **as a whole vector**. A row with $n_z$ zeros therefore lives on a unit hemisphere inside the orthogonal complement of its zero-partner rows, a sphere of dimension $d = i - 1 - n_z$. Sample there and the constraint holds by construction. No proposal can ever be infeasible, so the divergence mechanism of Family 1 simply does not exist. The joint delta over a row's zeros contributes $-\tfrac{1}{2}\log\det(A'A)$, where the columns of $A$ are the zero-partner rows. This is Axen's "divide by the determinant of the R factor."

My first two implementations (`qr`, `qr2`) followed the recipe literally: Householder QR of $A$, complement basis from the trailing columns of the full $Q$, hemisphere chart via a gnomonic map or $\tanh$ sticks. They achieved exactly zero divergences, and about **40% of the ESS** of the forced-zero family. Swapping the chart did nothing, which was the tell. The problem is the basis: Householder's complement columns are an arbitrary orthonormal frame that *spins* as the earlier rows move, so the raw coordinates keep changing meaning underneath the sampler.

## 4. A Canonical Triangular Basis

The fix is to anchor the basis to the matrix entries. For each free column $k$ of row $i$, take $e_k$ and fill in the row's forced columns by back-substitution through the zero constraints. This is exactly the Family 1 forced-entry fill, applied to *basis vectors* instead of the sampled row. Collect these as the columns of $T$. Splitting coordinates into forced ($Z$) and free ($F$) sets gives $T_F = I$ and $T_Z = -(A_Z')^{-1} A_F'$, and orthonormalization costs one small Cholesky factor $S = \operatorname{chol}(T'T)$: the map $v = T (S')^{-1} z$ satisfies $\lVert v \rVert = \lVert z \rVert$, so the sphere-measure bookkeeping goes through untouched.

Two identities then make the whole thing almost embarrassingly cheap.

1.  **The delta term factors.** The zero-partner rows have triangular support, so $A_Z$ is triangular with the partners' diagonals on its diagonal, and Sylvester's determinant identity gives $$
    \det(A'A) = \Big(\prod_{j \in Z} L_{jj}\Big)^{2} \det(T'T).
    $$ The conditioning term $-\tfrac{1}{2}\log\det(A'A)$ splits into the familiar Family 1 correction $-\sum_j \log L_{jj}$, which now falls out *analytically* with no linear algebra at all, plus $-\sum_k \log S_{kk}$, which rides along free with the orthonormalization Cholesky. I find it satisfying that the correction I originally derived as a hand-computed Jacobian in the forced-value construction reappears here as the determinant of the triangular block of a Gram factorization.

2.  **The Gram is small.** Since $T_F = I$, we have $T'T = I + W'W$ where $W$ holds only the $n_z$ forced-coordinate rows, so the crossprod works on length-$n_z$ vectors rather than length-$(i-1)$ ones.

Per constrained row, the entire construction is one $d \times d$ `cholesky_decompose` and some small matrix-vector products. No QR anywhere. I call this model `tri`; the full Stan code is in the appendix.

The subspace family has a second member: a "donut" variant, one extra parameter per row. Normalize a free vector, $u = y / \lVert y \rVert$, and the row's subsphere coordinates come along with beautifully isotropic geometry — except that the normalization map has a singularity at the origin, and the implicit chi-distributed radius that $-\tfrac{1}{2} y'y$ places on $\lVert y \rVert$ lets the sampler wander right into it. The cure is also Seth Axen's, from the *A Better Unit Vector* Discourse thread [@axen2022]: put an explicit prior on the radius that repels the norm from the origin. I use his Gaussian-centered form, replacing $-\tfrac{1}{2}\lVert y \rVert^2$ with $-\tfrac{1}{2}(\lVert y \rVert - m)^2$. The radial factor is independent of the direction, so the $\Omega$ posterior is untouched (verified against `carpenter`: max free-cell mean difference 0.0046); the only thing that changes is the geometry the sampler sees. Across the five $K = 7$, $\eta = 4$ patterns:

| radial prior | total divergences | median min ESS | median ESS / 1k gradients |
|:---|--:|--:|--:|
| $m = 0$ (implicit chi) | 2,549 | 4,778 | 19 |
| $m = 2$ | 349 | 8,106 | 80 |
| $m = 4$ | 8 | 10,778 | 192 |
| $m = 6$ | **0** | 9,045 | 162 |

The radial prior is worth two orders of magnitude in divergences and a factor of eight in gradient efficiency. Everything below runs the donut at $m = 6$. I still prefer `tri` as the default — no tuning knob and $K - 1$ fewer parameters — but the subspace family has two clean members.

## 5. The Bake-Off at K = 7

Setup: $K = 7$, $\eta = 4$, prior-only, five random zero patterns with roughly half of the 21 lower-triangular cells zeroed, 4 chains $\times$ 2000 draws after 1000 warmup, default `adapt_delta = 0.8`. Correctness was checked on a held-out pattern first: all twelve posteriors match the `carpenter` baseline within Monte Carlo error, and the constrained cells come back as numerically exact zeros ($\sim 10^{-19}$).

**Divergences** out of 8,000 post-warmup transitions:

| pattern | carpenter | pinkney | ldl | ldl2 | cvine | rowscale | ldl3 | schur | qr | qr2 | tri | donut |
|:--|--:|--:|--:|--:|--:|--:|--:|--:|--:|--:|--:|--:|
| seed 1 | 34 | 53 | 53 | 26 | 41 | 44 | 40 | 40 | 0 | 0 | **0** | 0 |
| seed 2 | 61 | 65 | 74 | 107 | 64 | 78 | 104 | 60 | 0 | 0 | **0** | 0 |
| seed 3 | 77 | 44 | 67 | 57 | 69 | 44 | 68 | 59 | 0 | 0 | **0** | 0 |
| seed 4 | 137 | 157 | 130 | 112 | 186 | 155 | 152 | 123 | 0 | 0 | **0** | 0 |
| seed 5 | 110 | 94 | 114 | 54 | 141 | 114 | 102 | 29 | 0 | 0 | **0** | 0 |
| **total** | 419 | 413 | 438 | 356 | 501 | 435 | 466 | 311 | **0** | **0** | **0** | **0** |

**Min bulk ESS** over the free cells of $\Omega$ (8,000 draws):

| pattern | carpenter | pinkney | ldl | ldl2 | cvine | rowscale | ldl3 | schur | qr | qr2 | tri | donut |
|:--|--:|--:|--:|--:|--:|--:|--:|--:|--:|--:|--:|--:|
| seed 1 | 11639 | 10566 | 9979 | 10618 | 10219 | 11319 | 9840 | 11036 | 3995 | 3362 | **11430** | 10676 |
| seed 2 | 10817 | 9866 | 10826 | 9772 | 10188 | 10153 | 10361 | 10040 | 4049 | 3728 | **11427** | 9045 |
| seed 3 | 9344 | 11014 | 10244 | 10314 | 10946 | 9420 | 10625 | 10787 | 4334 | 4304 | 9574 | 9429 |
| seed 4 | 8395 | 9159 | 8326 | 9257 | 8771 | 9463 | 8266 | 9458 | 3701 | 3815 | **9582** | 8161 |
| seed 5 | 10381 | 10010 | 9268 | 8760 | 10301 | 9218 | 9896 | 4770 | 4267 | 4153 | **11284** | 8619 |
| **median** | 10381 | 10010 | 9979 | 9772 | 10219 | 9463 | 9896 | 10040 | 4049 | 3815 | **11284** | 9045 |

**Efficiency** (medians over the five patterns):

|  | carpenter | pinkney | ldl | ldl2 | cvine | rowscale | ldl3 | schur | qr | qr2 | tri | donut |
|:--|--:|--:|--:|--:|--:|--:|--:|--:|--:|--:|--:|--:|
| min ESS / 1000 gradients | 188.5 | 179.6 | 185.9 | 182.2 | 188.9 | 176.3 | 182.1 | 182.0 | 75.0 | 72.5 | **207.3** | 161.8 |
| total wall clock, 5 fits (s) | 4.53 | 4.10 | 4.40 | 4.51 | 5.14 | 4.21 | 4.51 | 4.15 | 2.07 | 2.06 | **1.95** | 2.39 |

The `tri` column is the whole story: zero divergences on every pattern, the highest median ESS in the table, the best ESS per gradient, and the lowest wall clock. The subspace geometry means there is no boundary to fight, and the anchored basis means the sampler is not chasing a spinning frame. The usual robustness-versus-speed trade dissolves.

## 6. Does It Hold Up? Scaling K and $\eta$

A single $(K, \eta)$ corner is not a benchmark, so I reran a representative subset over $K \in \{5, 7, 9, 15, 25\}$ and $\eta \in \{1, 2, 3, 4\}$: six forced-zero models, the literal QR implementation `qr2`, `tri`, and the donut at $m = 6$. One Bernoulli(0.5) zero pattern per $K$ (realized: 6 of 10 cells at $K=5$ up to 143 of 300 at $K=25$), shared across $\eta$, same sampler settings as before. One fit per configuration, so the single-fit noise caveat from Section 2 applies to small gaps; the patterns below are not small.

```{r}
#| label: setup
#| warning: false
#| message: false

library(ggplot2)
library(dplyr)
library(tidyr)

source("../../R/theme_blog.R")

sweep <- read.csv("sweep-results.csv") |>
  mutate(
    family = case_when(
      model == "tri" ~ "Subspace (triangular basis)",
      model == "qr2" ~ "Subspace (QR basis)",
      model == "donut" ~ "Subspace (donut)",
      .default = "Forced zeros"
    ),
    div_pct = 100 * divergences / 8000,
    eta_lab = factor(paste0("eta == ", eta),
                     levels = paste0("eta == ", sort(unique(eta))))
  )

fam_colors <- c(
  "Forced zeros" = blog_colors$gold,
  "Subspace (QR basis)" = blog_colors$purple,
  "Subspace (donut)" = blog_colors$blue,
  "Subspace (triangular basis)" = blog_colors$teal
)
```

```{r}
#| label: fig-divergence-scaling
#| fig-cap: "Divergence rate by matrix dimension and LKJ concentration. Each thin gold line is one forced-zero parameterization; lines that stop mid-panel are models that failed to initialize at the next K. The teal line (tri) is exactly zero in every panel. The purple spike is qr2's catastrophic run at K = 15, eta = 2: one chain stuck at 100% divergence. The blue donut line keeps a residual sub-1.5% rate at eta = 1."
#| fig-height: 4.5
#| fig-width: 12
#| out-width: "100%"
#| warning: false

ggplot(sweep |> filter(!failed),
       aes(x = K, y = div_pct, group = model, color = family)) +
  geom_line(aes(linewidth = family, alpha = family)) +
  geom_point(size = 1.6) +
  facet_wrap(~eta_lab, nrow = 1, labeller = label_parsed) +
  scale_color_manual(values = fam_colors, name = NULL) +
  scale_linewidth_manual(values = c(0.6, 1.1, 1.1, 1.3), guide = "none") +
  scale_alpha_manual(values = c(0.65, 1, 1, 1), guide = "none") +
  scale_x_continuous(breaks = c(5, 7, 9, 15, 25)) +
  theme_blog() +
  labs(
    x = "K (matrix dimension)",
    y = "Divergent transitions (%)",
    title = "Divergences: Forced Zeros Degrade, the Triangular Basis Doesn't",
    subtitle = "One Bernoulli(0.5) zero pattern per K, 8,000 post-warmup transitions per fit; missing points failed to initialize"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-efficiency-scaling
#| fig-cap: "Sampling efficiency (min bulk ESS over free correlations per 1,000 gradient evaluations, log scale) by dimension and concentration."
#| fig-height: 4.5
#| fig-width: 12
#| out-width: "100%"
#| warning: false

ggplot(sweep |> filter(!failed),
       aes(x = K, y = ess_per_1k_grad, group = model, color = family)) +
  geom_line(aes(linewidth = family, alpha = family)) +
  geom_point(size = 1.6) +
  facet_wrap(~eta_lab, nrow = 1, labeller = label_parsed) +
  scale_color_manual(values = fam_colors, name = NULL) +
  scale_linewidth_manual(values = c(0.6, 1.1, 1.1, 1.3), guide = "none") +
  scale_alpha_manual(values = c(0.65, 1, 1, 1), guide = "none") +
  scale_x_continuous(breaks = c(5, 7, 9, 15, 25)) +
  scale_y_log10() +
  theme_blog() +
  labs(
    x = "K (matrix dimension)",
    y = "min ESS per 1,000 gradients (log)",
    title = "Efficiency: The Triangular Basis Wins at Every Size",
    subtitle = "Higher is better. The QR basis pays a growing geometry tax; the forced family drops out entirely at K = 25."
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: failed-fits
#| echo: false

sweep |>
  filter(failed) |>
  count(model, K) |>
  mutate(n = paste0(n, " of 4 etas")) |>
  pivot_wider(names_from = K, values_from = n, names_prefix = "K = ",
              values_fill = "—") |>
  knitr::kable(
    caption = "Outright failures: initialization could not find a feasible point after 100 attempts per chain."
  )
```

Three things jump out.

**The divergence tax explodes exactly where you would want to use this prior.** At $\eta = 1$, the uniform LKJ, the forced-zero family runs 13--17% divergent at $K = 5$, and by $K = 7$--$9$ it is at **85--94%**: five of every six transitions divergent. That is not a tuning problem; that is the uniform prior putting real mass right where the forced values overshoot the stick. (A fun artifact visible in the $\eta = 1$ panel: four of the gold lines coincide exactly. With the seed fixed, `carpenter`, `pinkney`, `ldl2`, and `ldl3` are *literally the same density in raw coordinates* — the correlation-scale bounds never bind, by Cauchy--Schwarz — so their chains agree bit-for-bit until floating-point chaos separates them. The distinct-looking K = 7 table entries in Section 5 are that chaos, not different geometry.)

**Initialization failure is the same boundary, expressed terminally.** `schur` cannot find a feasible starting point in 100 attempts per chain for this sweep's patterns from $K = 7$ up (it ran fine on all five Section 5 patterns at the same size; feasible initialization is pattern-dependent luck). `cvine` joins it at $K = 15$. At $K = 25$, with 143 forced cells, *none of the six forced models* can start at any $\eta$. This is the scaling fact that matters: the feasible set the forced construction must hit by chance gets exponentially thin as the pattern grows, and `adapt_delta` cannot rescue a model that never takes its first draw. The subspace models are immune by construction, since every point of their parameter space is feasible.

Since `tri` is the recommendation, here is its full scaling record on its own. Divergences are zero in every cell of this table, so what is left to watch is efficiency: min bulk ESS holds close to the 8,000-draw ceiling through $K = 15$ and comes down to a still-comfortable \~1,200 at $K = 25$, where the min is taken over 157 free correlations. ESS per gradient decays gently with $K$, is nearly flat in $\eta$, and the wall clock stays in seconds throughout.

```{r}
#| label: tri-scaling-table

tri_tab <- sweep |>
  filter(model == "tri") |>
  mutate(cell = sprintf("%s (%s)",
                        format(round(min_ess), big.mark = ","),
                        round(ess_per_1k_grad))) |>
  select(K, N_zero, eta, cell, seconds)

tri_wide <- tri_tab |>
  select(K, N_zero, eta, cell) |>
  pivot_wider(names_from = eta, values_from = cell,
              names_prefix = "eta = ") |>
  left_join(
    tri_tab |>
      group_by(K) |>
      summarize(`wall clock (s)` = sprintf("%.1f-%.1f",
                                           min(seconds), max(seconds)),
                .groups = "drop"),
    by = "K"
  ) |>
  mutate(`zero / free cells` = paste0(N_zero, " / ", choose(K, 2) - N_zero),
         .after = K) |>
  select(-N_zero)

knitr::kable(
  tri_wide, align = c("r", "c", rep("c", 4), "c"),
  caption = "tri across the full sweep: min bulk ESS of 8,000 draws (ESS per 1,000 gradients in parentheses). Divergences are 0 in every configuration; wall clock is the range across eta for 4 parallel chains."
)
```

**Anchoring the basis is not a nicety.** The literal QR implementation survives everywhere the forced family dies, but it pays for the spinning frame: at $K = 25$, $\eta = 1$ it needed 852 seconds against `tri`'s 6.8, and at $K = 15$, $\eta = 2$ it failed catastrophically, one chain pinned at 100% divergence and a min ESS of 7. (The forced family has its own catastrophic mode: `ldl3` at $K = 9$, $\eta = 2$ came back with a min ESS of 19.) Interestingly, `qr2`'s *raw* min ESS at $K = 25$ is higher than `tri`'s — about 3,900 vs 1,200 of 8,000 draws — but it burns 6--100$\times$ the gradient evaluations to get it. Per gradient, and per second, `tri` is the class of the field: the only model at exactly zero divergences in all twenty configurations (the donut keeps a residual sub-1.5% rate at $\eta = 1$), flat wall clock, and the best ESS per gradient in every cell but one ($K = 25$, $\eta = 4$, where the donut edges it).

## 7. When to Use Which

| Scenario | Recommendation |
|:---|:---|
| Any new model with structural zeros | `tri`: exact constraints, no divergence tax, no efficiency tax |
| Small $K$, sparse zeros, existing forced-zero code, deadlines | Family 1 is fine with `adapt_delta` raised and divergences monitored |
| $K \gtrsim 15$ with dense zero patterns | Family 1 may not even initialize; `tri` is the only option here that is both feasible and efficient |
| Huge $K$ | `tri`, but profile the per-row `cholesky_decompose`; the $d \times d$ factorizations are small but not free |
| Zeros plus *bounds* on the free correlations | Family 1 machinery composes naturally with bounded transforms [@pinkney2024; @pinkneytalk]; porting bounds into the subspace view is open |

## 8. Discussion

The key insight is geometric: **a constraint you enforce is a boundary; a constraint you parameterize away is not there at all**. The forced-zero family enforces, and pays on a sliding scale: a persistent divergence rate at small $K$, collapsing ESS at moderate $K$, and outright initialization failure once the pattern is dense enough that random starting points can never satisfy every forced cell at once. Axen's subspace construction [@axen2023] parameterizes away, and the only thing it was missing was a basis that respects the triangular structure of the problem. Once the complement basis is anchored to the matrix entries, the geometry cost vanishes and the exact method is simultaneously the safest, the most efficient per gradient, and the fastest on the wall clock.

The measurement lesson is worth repeating too. Identical densities implemented with different arithmetic produced divergence counts spanning a 30% range across five patterns. Single-fit divergence comparisons between parameterizations are mostly noise unless the gap is qualitative, like several hundred versus exactly zero.

Two threads are left open. Bounded correlations plus zeros in the subspace view would need the feasible region of a row intersected with a box, which no longer factors as a sphere. And the per-row Cholesky factorizations, while small, deserve profiling at $K$ in the hundreds. Both feel tractable.

## Acknowledgments

The exact-constraint subspace idea, zeros as orthogonality constraints with the sample lifted from a lower-dimensional hemisphere through the orthogonal complement, including the $\det R$ Jacobian factor, is due to Seth Axen [@axen2023], whose radial prior [@axen2022] the donut variant relies on. The `tri` model is that idea with a canonical triangular choice of complement basis and the determinant factored analytically. Thanks to Bob Carpenter for the compact zeros-only formulation benchmarked here as `carpenter`, and for the Discourse conversation that prompted this bake-off. The C-vine construction rests on the vine and partial-correlation results of @joe2006 and @lewandowski2009.

## Appendix: Stan Code for `tri`

``` {.stan filename="unitvec-tri-corr-zeros.stan"}
{{< include unitvec-tri-corr-zeros.stan >}}
```
