---
title: "Structural Zeros in Correlation Matrices: A Twelve-Parameterization Bake-Off"
author: "Sean Pinkney"
date: "2026-07-15"
date-modified: "2026-07-25"
categories: [stan, correlation, parameterizations]
bibliography: references.bib
format:
  html:
    code-fold: true
---

## 1. The Problem with Zeros in Correlation Matrices

Sometimes you know, before seeing any data, that certain correlations are exactly zero. For example, a priori knowledge that two variables are *marginally*^[[\@ethothepie.bsky.social](https://bsky.app/profile/ethothepie.bsky.social) corrected the original here where I incorrectly implied the zeros in the correlation matrix are conditionally unrelated when this is marginal] unrelated occurs in graphical models as either an explicit assumption or by pre-existing knowledge that rules out the variable coupling. Building this structure directly through the correlation matrix parameterization is what this blog post is about.

Sampling from an LKJ prior and setting selected cells to zero does not work because the resulting matrix will generally not be positive definite. Instead, the zeros have to be part of the parameterization. In this post I compare twelve Stan [@carpenter2017] implementations that target the same distribution. I check that they agree numerically, then compare divergences, effective sample size, gradient evaluations, and run time.

The implementations fall into two groups. The first solves for individual Cholesky entries so that the requested correlations are zero. The second samples each row in the subspace satisfying all of its zero constraints. I also introduce a triangular basis for the second approach. In these examples it removes the divergences seen in the entry-wise construction without the loss of efficiency I found with a QR basis.

## 2. Family 1: Forced Zeros

The first group uses the constrained-Cholesky approach I introduced in @pinkney2024 and presented at StanCon [@pinkneytalk]. The original method also allows a correlation to be fixed at a nonzero value or restricted to an interval. Writing $\Omega = L L'$ with unit-norm rows of $L$, the correlation in cell $(i,j)$ is $$
\Omega_{ij} = b_1 + L_{ij} L_{jj}, \qquad
b_1 = \sum_{k<j} L_{ik} L_{jk},
$$ This is affine in $L_{ij}$, so setting the correlation to zero gives $L_{ij}=-b_1/L_{jj}$ and uses no parameter. A bound on $\Omega_{ij}$ similarly becomes a bound on $L_{ij}$. Conditioning on $\Omega_{ij}=0$ contributes $-\log L_{jj}$ to the log density for each zero. The `carpenter` variant below is Bob Carpenter's compact formulation of this case from a [Stan Discourse discussion](https://discourse.mc-stan.org/t/new-users-guide-intro-sem-and-user-defined-constrained-parameters-chapters/41430):

``` stan
if (is_zero(i, j, zeros, zero_idx)) {
  real b = dot_product(L[j, 1:(j - 1)], L[i, 1:(j - 1)]);
  L[i, j] = -b / L[j, j];   // implies Omega[i, j] == 0
  jacobian += -log(L[j, j]);
  zero_idx += 1;
}
```

The free entries can be parameterized in several ways. I compare the following eight variants:

| label | free-cell parameterization |
|:-----------------------------------|:-----------------------------------|
| `carpenter` | direct Cholesky entries, bounded by the remaining stick |
| `pinkney` | correlation-scale value divided by $L_{jj}$, multiplicative stick [@pinkney2024; @pinkneytalk] |
| `ldl` / `ldl2` / `ldl3` | $LDL'$ factorization [@pinkney2024]; stick tracked via `log_diff_exp`, a `log1m` ratio, or the natural scale |
| `cvine` | C-vine partial correlations with the LKJ Beta factorization [@joe2006; @lewandowski2009] |
| `rowscale` | direct Cholesky entries, row-scaling stick |
| `schur` | canonical partial correlations, $L_{ij} = r \tanh(x)$ |

These variants share the same constraint. The value $-b_1/L_{jj}$ is computed without reference to the remaining row norm, $\sqrt{1 - \sum_{k<j} L_{ik}^2}$, and can exceed the available range. A later square root then receives a negative argument. In the $K=7$ benchmarks below, each variant has divergences in about 1% of post-warmup transitions.

The three LDL variants are also useful for judging Monte Carlo variability. They define the same density and differ only in how some floating-point operations are arranged, but their total divergences over five patterns range from 356 to 466. Differences of that size should not be interpreted as evidence that one parameterization is better from a single set of fits.

## 3. Family 2: Exact-Constraint Subspaces

The second approach comes from Seth Axen's Turing.jl discussion of positive definite matrices with structural zeros [@axen2023]:

> For each column j, each column i for i \< j that is orthogonal to column j decreases the dimensionality of the hemisphere by 1. So we simply construct a point on a lower dimensional hemisphere and use the QR decomposition of the orthogonal columns to lift it to a higher dimensional hemisphere.

Row $i$ of a correlation Cholesky factor is a unit vector with a positive diagonal element. Since row $j$ is zero after column $j$, the condition $\Omega_{ij}=0$ is equivalent to requiring rows $i$ and $j$ to be orthogonal. If row $i$ has $n_z$ constrained correlations, it can be sampled on the unit hemisphere in the orthogonal complement of those $n_z$ rows. The dimension of this hemisphere is $d=i-1-n_z$. The joint conditioning term for the row is $-\tfrac12\log\det(A'A)$, where the columns of $A$ are the rows to which it must be orthogonal.

My first two implementations, `qr` and `qr2`, construct this complement with a Householder QR decomposition. They use either a gnomonic map or the hyper-tanh peel [@pinkney2026] for the hemisphere coordinates. Both models have zero divergences in the $K=7$ benchmark, but their minimum ESS is about 40% of the entry-wise models. Changing the hemisphere map has little effect. The larger difference comes from the complement basis: the trailing columns of $Q$ change as the preceding rows change, so the interpretation of the unconstrained coordinates also changes during sampling.

## 4. A Canonical Triangular Basis

I instead construct a basis tied to the free matrix entries. For each free column $k$ of row $i$, I start with the coordinate vector $e_k$ and solve for the constrained columns by back-substitution. This is the entry-wise construction from the previous section applied to a basis vector rather than to the sampled row. If $Z$ and $F$ denote the constrained and free columns, the resulting basis matrix $T$ satisfies $T_F=I$ and $T_Z=-(A_Z')^{-1}A_F'$.

To orthonormalize this basis, let $S=\operatorname{chol}(T'T)$. Then $v=T(S')^{-1}z$ preserves the norm of $z$. I parameterize $z$ using the row transformation from @pinkney2025. With $t=y'y$, $$
z = \frac{\sqrt{t+2}}{t+1}y, \qquad L_{ii}=\frac{1}{t+1},
$$ and the log Jacobian is $\tfrac12(d-2)\log(t+2)-(d+1)\log(1+t)$.

Two identities reduce the amount of linear algebra required.

1.  The zero-partner rows have triangular support, so $A_Z$ is triangular with the partners' diagonals on its diagonal. Sylvester's determinant identity gives $$
    \det(A'A) = \Big(\prod_{j \in Z} L_{jj}\Big)^{2} \det(T'T).
    $$ Thus $-\tfrac12\log\det(A'A)$ becomes $-\sum_j\log L_{jj}-\sum_k\log S_{kk}$. The second term is available from the Cholesky factor already used to orthonormalize $T$.

2.  Since $T_F=I$, $T'T=I+W'W$, where $W$ contains only the $n_z$ constrained rows. The cross-product therefore uses vectors of length $n_z$ rather than $i-1$.

Each constrained row requires one $d\times d$ `cholesky_decompose` and several matrix-vector products. I call this model `tri`; the Stan code is in the appendix.

I also test an over-parameterized version, called `donut`, that uses one additional parameter per row. It normalizes a free vector, $u=y/\lVert y\rVert$, before mapping it into the constrained subspace. The normalization is rotationally symmetric but singular at the origin. Following @axen2022, I replace the usual radial term $-\tfrac12\lVert y\rVert^2$ with $-\tfrac12(\lVert y\rVert-m)^2$ so that the typical radius is away from zero. Since this term only changes the radius, it does not change the induced distribution on $\Omega$. The largest difference in a free-cell posterior mean relative to `carpenter` is 0.0046. Across the five $K=7$, $\eta=4$ patterns:

| radial prior | total divergences | median min ESS | median ESS / 1k gradients |
|:-------------------|-----------------:|-----------------:|-----------------:|
| $m = 0$ (implicit chi) | 2,549 | 4,778 | 19 |
| $m = 2$ | 349 | 8,106 | 80 |
| $m = 4$ | 8 | 10,778 | 192 |
| $m = 6$ | **0** | 9,045 | 162 |

Increasing $m$ removes the divergences and improves efficiency up to $m=4$; $m=6$ gives zero divergences in these runs. I use $m=6$ below. The `tri` model does not require this additional choice and uses $K-1$ fewer parameters.

## 5. The Bake-Off at K = 7

Setup: $K = 7$, $\eta = 4$, prior-only, five random zero patterns with roughly half of the 21 lower-triangular cells zeroed, 4 chains $\times$ 2000 draws after 1000 warmup, default `adapt_delta = 0.8`. Correctness was checked on a held-out pattern first: all twelve posteriors match the `carpenter` baseline within Monte Carlo error, and the constrained cells come back as numerically exact zeros ($\sim 10^{-19}$).

**Divergences** out of 8,000 post-warmup transitions:

| pattern | carpenter | pinkney | ldl | ldl2 | cvine | rowscale | ldl3 | schur | qr | qr2 | tri | donut |
|:-----|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|
| seed 1 | 34 | 53 | 53 | 26 | 41 | 44 | 40 | 40 | 0 | 0 | **0** | 0 |
| seed 2 | 61 | 65 | 74 | 107 | 64 | 78 | 104 | 60 | 0 | 0 | **0** | 0 |
| seed 3 | 77 | 44 | 67 | 57 | 69 | 44 | 68 | 59 | 0 | 0 | **0** | 0 |
| seed 4 | 137 | 157 | 130 | 112 | 186 | 155 | 152 | 123 | 0 | 0 | **0** | 0 |
| seed 5 | 110 | 94 | 114 | 54 | 141 | 114 | 102 | 29 | 0 | 0 | **0** | 0 |
| **total** | 419 | 413 | 438 | 356 | 501 | 435 | 466 | 311 | **0** | **0** | **0** | **0** |

**Min bulk ESS** over the free cells of $\Omega$ (8,000 draws):

| pattern | carpenter | pinkney | ldl | ldl2 | cvine | rowscale | ldl3 | schur | qr | qr2 | tri | donut |
|:-----|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|
| seed 1 | 11639 | 10566 | 9979 | 10618 | 10219 | 11319 | 9840 | 11036 | 3995 | 3362 | **11238** | 10676 |
| seed 2 | 10817 | 9866 | 10826 | 9772 | 10188 | 10153 | 10361 | 10040 | 4049 | 3728 | **11923** | 9045 |
| seed 3 | 9344 | 11014 | 10244 | 10314 | 10946 | 9420 | 10625 | 10787 | 4334 | 4304 | 10655 | 9429 |
| seed 4 | 8395 | 9159 | 8326 | 9257 | 8771 | 9463 | 8266 | 9458 | 3701 | 3815 | **9400** | 8161 |
| seed 5 | 10381 | 10010 | 9268 | 8760 | 10301 | 9218 | 9896 | 4770 | 4267 | 4153 | **11845** | 8619 |
| **median** | 10381 | 10010 | 9979 | 9772 | 10219 | 9463 | 9896 | 10040 | 4049 | 3815 | **11238** | 9045 |

**Efficiency** (medians over the five patterns):

|   | carpenter | pinkney | ldl | ldl2 | cvine | rowscale | ldl3 | schur | qr | qr2 | tri | donut |
|:-----|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|
| min ESS / 1000 gradients | 188.5 | 179.6 | 185.9 | 182.2 | 188.9 | 176.3 | 182.1 | 182.0 | 75.0 | 72.5 | **212.3** | 161.8 |
| total wall clock, 5 fits (s) | 4.53 | 4.10 | 4.40 | 4.51 | 5.14 | 4.21 | 4.51 | 4.15 | 2.07 | 2.06 | **1.86** | 2.39 |

The `tri` model has zero divergences for all five patterns. It also has the highest median minimum ESS, the highest median ESS per 1,000 gradients, and the shortest total run time. The QR models also remove divergences, but their changing complement basis produces substantially lower ESS. The triangular basis avoids that loss in this benchmark.

## 6. Does It Hold Up? Scaling K and $\eta$

I next repeat the comparison for $K\in\{5,7,9,15,25\}$ and $\eta\in\{1,2,3,4\}$. This sweep includes six entry-wise models, `qr2`, `tri`, and the donut model with $m=6$. For each $K$, I generate one Bernoulli(0.5) zero pattern and reuse it for all four values of $\eta$. The realized number of zeros ranges from 6 of 10 cells at $K=5$ to 143 of 300 at $K=25$. Sampler settings are unchanged. There is one fit per configuration, so small differences should be interpreted with the variability from Section 2 in mind.

```{r}
#| label: setup
#| warning: false
#| message: false

library(ggplot2)
library(dplyr)
library(tidyr)

source("../../R/theme_blog.R")

sweep <- read.csv("sweep-results.csv") |>
  mutate(
    family = case_when(
      model == "tri" ~ "Subspace (triangular basis)",
      model == "qr2" ~ "Subspace (QR basis)",
      model == "donut" ~ "Subspace (donut)",
      .default = "Forced zeros"
    ),
    div_pct = 100 * divergences / 8000,
    eta_lab = factor(paste0("eta == ", eta),
                     levels = paste0("eta == ", sort(unique(eta))))
  )

fam_colors <- c(
  "Forced zeros" = blog_colors$gold,
  "Subspace (QR basis)" = blog_colors$purple,
  "Subspace (donut)" = blog_colors$blue,
  "Subspace (triangular basis)" = blog_colors$teal
)
```

```{r}
#| label: fig-divergence-scaling
#| fig-cap: "Divergence rate by matrix dimension and LKJ concentration. Each thin gold line is one entry-wise parameterization; missing points indicate initialization failures. The tri model has zero divergences in every panel. For qr2 at K = 15 and eta = 2, one chain has 100% divergences. The donut model remains below 1.5% at eta = 1 and has zero divergences otherwise."
#| fig-height: 4.5
#| fig-width: 12
#| out-width: "100%"
#| warning: false

ggplot(sweep |> filter(!failed),
       aes(x = K, y = div_pct, group = model, color = family)) +
  geom_line(aes(linewidth = family, alpha = family, linetype = family)) +
  geom_point(size = 1.6) +
  facet_wrap(~eta_lab, nrow = 1, labeller = label_parsed) +
  scale_color_manual(values = fam_colors, name = NULL) +
  scale_linewidth_manual(guide = "none", values = c(
    "Forced zeros" = 0.6, "Subspace (QR basis)" = 1.1,
    "Subspace (donut)" = 1.1, "Subspace (triangular basis)" = 1.3)) +
  scale_alpha_manual(guide = "none", values = c(
    "Forced zeros" = 0.65, "Subspace (QR basis)" = 1,
    "Subspace (donut)" = 1, "Subspace (triangular basis)" = 1)) +
  scale_linetype_manual(guide = "none", values = c(
    "Forced zeros" = "solid", "Subspace (QR basis)" = "solid",
    "Subspace (donut)" = "22", "Subspace (triangular basis)" = "solid")) +
  scale_x_continuous(breaks = c(5, 7, 9, 15, 25)) +
  theme_blog() +
  labs(
    x = "K (matrix dimension)",
    y = "Divergent transitions (%)",
    title = "Divergence Rate by Matrix Dimension",
    subtitle = "One Bernoulli(0.5) zero pattern per K and 8,000 post-warmup transitions per fit"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-efficiency-scaling
#| fig-cap: "Sampling efficiency (min bulk ESS over free correlations per 1,000 gradient evaluations, log scale) by dimension and concentration."
#| fig-height: 4.5
#| fig-width: 12
#| out-width: "100%"
#| warning: false

ggplot(sweep |> filter(!failed),
       aes(x = K, y = ess_per_1k_grad, group = model, color = family)) +
  geom_line(aes(linewidth = family, alpha = family, linetype = family)) +
  geom_point(size = 1.6) +
  facet_wrap(~eta_lab, nrow = 1, labeller = label_parsed) +
  scale_color_manual(values = fam_colors, name = NULL) +
  scale_linewidth_manual(guide = "none", values = c(
    "Forced zeros" = 0.6, "Subspace (QR basis)" = 1.1,
    "Subspace (donut)" = 1.1, "Subspace (triangular basis)" = 1.3)) +
  scale_alpha_manual(guide = "none", values = c(
    "Forced zeros" = 0.65, "Subspace (QR basis)" = 1,
    "Subspace (donut)" = 1, "Subspace (triangular basis)" = 1)) +
  scale_linetype_manual(guide = "none", values = c(
    "Forced zeros" = "solid", "Subspace (QR basis)" = "solid",
    "Subspace (donut)" = "22", "Subspace (triangular basis)" = "solid")) +
  scale_x_continuous(breaks = c(5, 7, 9, 15, 25)) +
  scale_y_log10() +
  theme_blog() +
  labs(
    x = "K (matrix dimension)",
    y = "min ESS per 1,000 gradients (log)",
    title = "Sampling Efficiency by Matrix Dimension",
    subtitle = "Higher is better; missing points indicate initialization failures"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: failed-fits
#| echo: false

sweep |>
  filter(failed) |>
  count(model, K) |>
  mutate(n = paste0(n, " of 4 etas")) |>
  pivot_wider(names_from = K, values_from = n, names_prefix = "K = ",
              values_fill = "—") |>
  knitr::kable(
    caption = "Outright failures: initialization could not find a feasible point after 100 attempts per chain."
  )
```

At $\eta=1$, the entry-wise models have 13--17% divergences for $K=5$ and 85--94% for $K=7$ and $K=9$. The uniform LKJ prior places more mass near the part of the parameter space where a solved Cholesky entry can exceed the remaining row norm. Four curves coincide in the $\eta=1$ panel because `carpenter`, `pinkney`, `ldl2`, and `ldl3` have the same density in their raw coordinates until floating-point differences separate their trajectories.

Initialization also becomes difficult for the entry-wise models. For these particular patterns, `schur` fails to initialize from $K=7$ onward and `cvine` fails at $K=15$. At $K=25$, none of the six entry-wise models finds a feasible initial value in 100 attempts per chain. This depends on the zero pattern—`schur` initialized for every pattern in Section 5—but the problem becomes more likely as the number of constraints increases. Every initial value in the subspace models satisfies the zero constraints by construction.

The `tri` model has zero divergences in all 20 configurations. The donut model has a small number at $\eta=1$ and none for $\eta\ge2$. Through $K=15$, the minimum bulk ESS for `tri` remains near the number of posterior draws. At $K=25$, the minimum is taken over 157 free correlations and ranges from about 5,100 to 6,600 for `tri`. The donut model ranges from about 4,700 to 6,600. Both complete in seconds.

```{r}
#| label: subspace-scaling-table

sub_tab <- sweep |>
  filter(model %in% c("tri", "donut")) |>
  mutate(cell = sprintf("%s (%s)%s",
                        format(round(min_ess), big.mark = ","),
                        round(ess_per_1k_grad),
                        ifelse(divergences > 0,
                               paste0(" [", divergences, " div]"), "")))

sub_wide <- sub_tab |>
  select(model, K, N_zero, eta, cell) |>
  pivot_wider(names_from = eta, values_from = cell,
              names_prefix = "eta = ") |>
  left_join(
    sub_tab |>
      group_by(model, K) |>
      summarize(`wall clock (s)` = sprintf("%.1f-%.1f",
                                           min(seconds), max(seconds)),
                .groups = "drop"),
    by = c("model", "K")
  ) |>
  mutate(`zero / free cells` = paste0(N_zero, " / ", choose(K, 2) - N_zero),
         .after = K) |>
  select(-N_zero) |>
  arrange(K, desc(model))

knitr::kable(
  sub_wide, align = c("l", "r", "c", rep("c", 4), "c"),
  caption = "tri and the donut (m_radius = 6) across the full sweep: min bulk ESS of 8,000 draws (ESS per 1,000 gradients in parentheses; divergent transitions in brackets where nonzero -- tri has none anywhere). Wall clock is the range across eta for 4 parallel chains."
)
```

The QR basis remains feasible when the entry-wise models fail, but it becomes expensive as $K$ grows. At $K=25$, $\eta=1$, `qr2` takes 852 seconds compared with 8.9 seconds for `tri`. At $K=15$, $\eta=2$, one `qr2` chain has 100% divergences and the minimum ESS is 7. The `ldl3` model has a similar isolated failure at $K=9$, $\eta=2$, with a minimum ESS of 19. At $K=25$, the minimum ESS for `qr2` is about 3,500--4,100, compared with 5,100--6,600 for `tri`, and `qr2` requires many more gradient evaluations. `tri` has the highest ESS per gradient in 17 of the 20 configurations; `qr2` is higher in three of the $K=5$ configurations.

## 7. When to Use Which

| Scenario | Recommendation |
|:-----------------------------------|:-----------------------------------|
| Structural zeros only | Use `tri` |
| Structural zeros at very large $K$ | Use `tri`, but profile the per-row $d\times d$ Cholesky factorizations |
| Zeros plus other known correlation values | Use the entry-wise method [@pinkney2024; @pinkneytalk]; `tri` currently handles zeros only |
| Zeros plus bounds on other correlations | Use the entry-wise method, which already supports bounded transforms [@pinkney2024; @pinkneytalk] |

For structural zeros alone, I do not see a reason to prefer one of the entry-wise parameterizations. In these benchmarks `tri` is more reliable and at least as efficient. The entry-wise method remains useful because it handles more general constraints: correlations fixed at nonzero values and bounds on otherwise free correlations.

## 8. Discussion

The entry-wise construction solves each zero constraint but leaves the boundary of the feasible row norm in the parameter space. This produces divergences for the smaller examples and initialization failures for the larger ones. The subspace construction removes that boundary because every parameter value maps to a row satisfying the constraints.

The choice of basis is important. A complement obtained directly from QR is valid, but its coordinates change with the earlier rows and sampling becomes less efficient. The triangular basis fixes the free coordinates to matrix entries and only solves for the constrained coordinates. In these benchmarks, that gives zero divergences without sacrificing effective sample size or run time.

There are two limitations I have not addressed here. Adding bounds on otherwise free correlations changes the row's feasible region from a sphere to the intersection of a sphere and a box. Also, the small Cholesky factorization performed for each constrained row should be profiled for matrices with dimensions in the hundreds.

## Acknowledgments

The exact-constraint subspace idea, zeros as orthogonality constraints with the sample lifted from a lower-dimensional hemisphere through the orthogonal complement, including the $\det R$ Jacobian factor, is due to Seth Axen [@axen2023], whose radial prior [@axen2022] the donut variant relies on. The `tri` model is that idea with a canonical triangular choice of complement basis and the determinant factored analytically. Thanks to Bob Carpenter for the compact zeros-only formulation benchmarked here as `carpenter`, and for the Discourse conversation that prompted this bake-off. The C-vine construction rests on the vine and partial-correlation results of @joe2006 and @lewandowski2009.

## Appendix: Stan Code for All Twelve Parameterizations

::: panel-tabset
### tri

``` {.stan filename="unitvec-stereo-corr-zeros.stan"}
{{< include unitvec-stereo-corr-zeros.stan >}}
```

### donut

``` {.stan filename="unitvec-donut-corr-zeros.stan"}
{{< include unitvec-donut-corr-zeros.stan >}}
```

### carpenter

``` {.stan filename="carpenter-corr-zeros.stan"}
{{< include carpenter-corr-zeros.stan >}}
```

### pinkney

``` {.stan filename="pinkney-corr-zeros2.stan"}
{{< include pinkney-corr-zeros2.stan >}}
```

### ldl

``` {.stan filename="pinkney-corr-zeros-ldl.stan"}
{{< include pinkney-corr-zeros-ldl.stan >}}
```

### ldl2

``` {.stan filename="pinkney-corr-zeros-ldl2.stan"}
{{< include pinkney-corr-zeros-ldl2.stan >}}
```

### ldl3

``` {.stan filename="pinkney-corr-zeros-ldl3.stan"}
{{< include pinkney-corr-zeros-ldl3.stan >}}
```

### cvine

``` {.stan filename="cvine-corr-zeros.stan"}
{{< include cvine-corr-zeros.stan >}}
```

### rowscale

``` {.stan filename="pinkney-corr-zeros-rowscale.stan"}
{{< include pinkney-corr-zeros-rowscale.stan >}}
```

### schur

``` {.stan filename="pinkney-corr-zeros-schur.stan"}
{{< include pinkney-corr-zeros-schur.stan >}}
```

### qr

``` {.stan filename="qr-corr-zeros.stan"}
{{< include qr-corr-zeros.stan >}}
```

### qr2

``` {.stan filename="qr2-corr-zeros.stan"}
{{< include qr2-corr-zeros.stan >}}
```
:::
