---
title: "Block-Level LKJ Priors for Structured Correlation Matrices"
author: "Sean Pinkney"
date: "2026-03-31"
categories: [stan, correlation]
bibliography: references.bib
format:
  html:
    code-fold: true
---

## 1. The Problem with Priors on Large Correlation Matrices

The LKJ distribution [@lewandowski2009] is the default prior for correlation matrices in Stan [@carpenter2017]. For a $P \times P$ correlation matrix $R$, it has density: $$
p(R \mid \eta) \propto |R|^{\eta - 1}
$$ When $\eta = 1$ the prior is uniform over correlation matrices; when $\eta > 1$ it concentrates toward the identity. This works well for small, unstructured problems.

But many multivariate datasets have **known structure**. Consider $N$ subjects each measured at $M$ time points, giving $P = NM$ total dimensions. The correlation matrix is not an arbitrary $P \times P$ matrix as it has a natural decomposition into within-subject temporal correlations and between-subject cross-correlations. Slapping a single LKJ prior on the full matrix creates three problems:

1.  **Uniform shrinkage**: LKJ treats every off-diagonal entry the same. It penalizes strong within-subject temporal correlation just as hard as strong cross-subject correlation, even though these have completely different interpretations and plausible ranges.

2.  **Dimension-dependent concentration**: LKJ($\eta$) concentrates harder as $P$ grows. Adding more subjects (increasing $N$) makes the prior shrink all correlations toward zero.

3.  **No elicitation lever**: A practitioner might know that temporal autocorrelation is strong while cross-subject coupling is weak. With a single $\eta$ parameter for the full matrix, there is no way to encode this.

In this post I place LKJ priors on the **building blocks** of the correlation matrix instead of the assembled matrix. The compound-symmetric block-Toeplitz structure makes this decomposition tractable and the PACF parameterization makes it computationally clean.

This matters most in the regime where data are genuinely scarce. In longitudinal, multilevel, and panel settings, it is common to have fewer than 15 independent observational units: pilot studies, rare-disease cohorts, small classroom or clinic samples, intensive repeated-measures designs, and expensive laboratory experiments all produce this pattern. In those settings the number of correlation parameters implied by the full matrix can easily dwarf the amount of information in the data. That is exactly when we need to exploit structure rather than pretend the correlation matrix is unstructured.

## 2. The Compound-Symmetric Block-Toeplitz Structure

### 2.1 Definition

The model is easiest to define through two Toeplitz covariance components: $$
\Sigma_A = \sigma_A^2 T_A, \qquad
\Sigma_B = \sigma_B^2 T_B,
$$ where $T_A$ and $T_B$ are $M \times M$ Toeplitz correlation matrices and $$
\sigma_A^2 = s \cdot \frac{N}{N - 1}, \qquad \sigma_B^2 = N(1-s),
$$ with a single mixing parameter $s \in (0, 1)$.

These two components induce the full $P \times P$ block matrix through the mean/deviation decomposition used in the likelihood. The diagonal blocks are $$
R_{\text{on}} = \frac{(N-1)\Sigma_A + \Sigma_B}{N}
= s T_A + (1-s) T_B,
$$ and the off-diagonal blocks are $$
R_{\text{off}} = \frac{\Sigma_B - \Sigma_A}{N}
= (1-s) T_B - \frac{s}{N-1} T_A.
$$ Equivalently, the full matrix can be written as $$
R = I_N \otimes R_{\text{on}} + (J_N - I_N) \otimes R_{\text{off}},
$$ where $J_N = \mathbf{1}_N \mathbf{1}_N^\top$.

This is the exact structure implemented in the Stan model. The key simplification is that every off-diagonal block is determined by the same two Toeplitz pieces and the same scalar $s$, so we never parameterize each block separately.

Interpretation:

-   $T_A$ governs the **block-specific** temporal component
-   $T_B$ governs the **shared across-block** temporal component
-   $s$ controls the balance: high $s$ emphasizes block-specific structure and weakens cross-block dependence; low $s$ increases the shared component

### 2.2 PACF Parameterization

A Toeplitz correlation matrix is positive definite if and only if all its partial autocorrelation (PACF) coefficients $\psi_1, \ldots, \psi_{M-1} \in (-1, 1)$ [@barndorff1973]. I map unconstrained parameters $\psi_{\text{raw}} \in \mathbb{R}^D$ to this interval via a scaled $\tanh$: $$
\psi_k = \tanh\!\left(\frac{\psi_{\text{raw}, k}}{\sqrt{D - 1}}\right)
$$ then recover the autocorrelation function (ACF) using the Durbin-Levinson recursion [@durbinlevinson; @levinson1946]. This guarantees positive definiteness by construction.

## 3. Block-Level LKJ Priors

### 3.1 The Standard (Full-Matrix) Approach

The standard LKJ prior on the full compound-symmetric matrix uses the log-determinant of the assembled $P \times P$ matrix: $$
\log p_{\text{full}}(\psi_A, \psi_B, s \mid \eta) = (\eta - 1) \log|R(\psi_A, \psi_B, s)| + \log|J_{\text{CS}}|
$$ where $|J_{\text{CS}}|$ is the Jacobian from the compound-symmetric parameterization. The full log-determinant decomposes as: $$
\log|R| = (N-1)\Big(M \log \tfrac{sN}{N-1} + \log|T_A|\Big) + \Big(M \log N(1-s) + \log|T_B|\Big)
$$

The problem is visible in this formula: the prior penalty on $T_A$ is amplified by $(N-1)(\eta - 1)$. **Adding subjects inflates the prior shrinkage on the block-specific temporal autocorrelation parameters in** $T_A$, even though the underlying temporal structure hasn't changed at all. By contrast, the $T_B$ term does not pick up the same $(N-1)$ amplification.

### 3.2 The Innovation: Priors on the Blocks

Instead of a single LKJ on $R$, we place **independent LKJ priors directly on** $T_A$ and $T_B$: $$
\log p_{\text{block}}(\psi_A, \psi_B \mid \eta_A, \eta_B) = (\eta_A - 1) \log|T_A(\psi_A)| + \log|J_A| + (\eta_B - 1) \log|T_B(\psi_B)| + \log|J_B|
$$ where $|J_A|$ and $|J_B|$ are the Jacobians for each Toeplitz block's PACF-to-correlation transformation.

This gives us three advantages:

1.  **Independent control**: Set $\eta_A = 4$ (stronger shrinkage on the block-specific component) while using $\eta_B = 2$ (milder shrinkage on the shared component), or any other combination.

2.  **Invariant to** $N$: The prior on $T_A$ depends only on the $M \times M$ block, not on how many subjects there are. Adding subjects brings more data, not more prior shrinkage on the block-specific Toeplitz component.

3.  **Interpretable elicitation**: "I believe the block-specific component is weak but the shared temporal component can persist" translates directly to $\eta_A > \eta_B$.

### 3.3 Why the Decomposition Works

The PACF parameterization makes this possible because:

-   Each Toeplitz block is parameterized by an **independent** PACF vector
-   The Jacobian of the PACF $\to$ Toeplitz transformation factorizes across blocks
-   The log-determinant of the full matrix decomposes into per-block terms (as shown above)

So the change-of-variables formula applies cleanly to each component, giving us a well-defined density on the building blocks without needing to reason about the assembled $P \times P$ matrix at all.

## 4. Efficient Likelihood

The compound-symmetric structure also enables an efficient likelihood. The quadratic form decomposes into $M \times M$ Toeplitz systems:

For each observation, define the block mean $\bar{\mathbf{z}} = \frac{1}{N}\sum_n \mathbf{z}^{(n)}$ and deviations $\tilde{\mathbf{z}}^{(n)} = \mathbf{z}^{(n)} - \bar{\mathbf{z}}$. Then: $$
\mathbf{z}^\top \Sigma^{-1} \mathbf{z} = N\,\bar{\mathbf{z}}^\top \Sigma_B^{-1} \bar{\mathbf{z}} + \sum_{n=1}^{N} \tilde{\mathbf{z}}^{(n)\top} \Sigma_A^{-1} \tilde{\mathbf{z}}^{(n)}
$$ where $\Sigma_A = \frac{sN}{N-1}T_A$ and $\Sigma_B = N(1-s)T_B$. Each Toeplitz solve uses the innovations algorithm [@brockwell2002] in $O(M^2)$ time. The total cost is $O(NM^2)$ per observation instead of $O(P^3) = O(N^3M^3)$.

## 5. Simulation Study

### 5.1 Setup

```{r}
#| label: setup
#| warning: false
#| message: false

library(cmdstanr)
library(mvtnorm)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(bayesplot)

source("../../R/theme_blog.R")
use_blog_bayesplot()
```

```{r}
#| label: compile
#| message: false
#| output: false

model <- cmdstan_model("compound-symmetric-block-toeplitz.stan")
```

```{r}
#| label: sim-functions

squash_pacf <- function(psi_raw) {
  D <- length(psi_raw)
  if (D <= 1) return(tanh(psi_raw))
  return(tanh(psi_raw / sqrt(D - 1)))
}

pacf_to_acf <- function(psi_raw) {
  D <- length(psi_raw)
  rho <- numeric(D + 1)
  rho[1] <- 1.0
  if (D == 0) return(rho)

  psi <- squash_pacf(psi_raw)
  phi <- numeric(D)

  for (k in 1:D) {
    psi_k <- psi[k]
    if (k > 1) {
      phi_prev <- phi[1:(k - 1)]
      for (j in 1:(k - 1)) {
        phi[j] <- phi_prev[j] - psi_k * phi_prev[k - j]
      }
    }
    phi[k] <- psi_k
    rho[k + 1] <- sum(phi[1:k] * rev(rho[1:k]))
  }

  return(rho)
}

build_toeplitz <- function(r) {
  n <- length(r)
  outer(seq_len(n), seq_len(n), function(i, j) r[abs(i - j) + 1])
}

build_full_R <- function(psiA, psiB, s, M, N) {
  rhoA <- pacf_to_acf(psiA)
  rhoB <- pacf_to_acf(psiB)

  sigmaA <- s * N / (N - 1)
  sigmaB <- (1 - s) * N
  a <- sigmaA * rhoA
  b <- sigmaB * rhoB

  r_on  <- ((N - 1) * a + b) / N
  r_off <- (b - a) / N

  T_on  <- build_toeplitz(r_on)
  T_off <- build_toeplitz(r_off)

  P <- M * N
  R <- matrix(0, P, P)
  for (i in 0:(N - 1))
    for (j in 0:(N - 1)) {
      idx_i <- (i * M + 1):((i + 1) * M)
      idx_j <- (j * M + 1):((j + 1) * M)
      R[idx_i, idx_j] <- if (i == j) T_on else T_off
    }
  R
}
```

### 5.2 Prior Predictive Comparison

Before fitting any data, let's see what the two priors imply about the lag-1 autocorrelation in $T_A$. To do this we can sample from the Stan model directly with no observations and the posterior is the prior, so these draws come from the exact prior induced by the model we will fit later.

```{r}
#| label: prior-densities
sample_prior_lag1 <- function(N_blocks, prior_on_full_corr, eta = 4, seed = 123) {
  K <- M * N_blocks

  fit_prior <- model$sample(
    data = list(
      M_blocksize = M,
      N_blocks = N_blocks,
      eta_lkj_a = eta,
      eta_lkj_b = eta,
      N = 0,
      K = K,
      y = matrix(numeric(0), nrow = 0, ncol = K),
      mu = rep(0, K),
      prior_on_full_corr = prior_on_full_corr
    ),
    chains = 1,
    parallel_chains = 1,
    iter_warmup = 500,
    iter_sampling = 1000,
    seed = seed,
    refresh = 0
  )

  draws_A <- fit_prior$draws("psi_A_raw", format = "matrix")
  lag1_A <- apply(draws_A, 1, function(x) pacf_to_acf(x)[2])

  data.frame(
    lag1 = lag1_A,
    Prior = if (prior_on_full_corr == 1) "Full-Matrix LKJ" else "Block-Level LKJ",
    N_blocks = factor(paste0("N[blocks] == ", N_blocks), levels = paste0("N[blocks] == ", c(2, 4, 8, 16)))
  )
}
```

```{r}
#| label: fig-prior-predictive
#| fig-cap: "Prior-implied distribution of the lag-1 ACF of T_A under block-level LKJ (stable across N_blocks) versus full-matrix LKJ (shrinks as N_blocks grows). Both use eta = 4 on T_A."
#| message: false
#| fig-height: 5
#| fig-width: 12
#| out-width: "100%"

set.seed(42)
M <- 5
eta <- 4

prior_df <- dplyr::bind_rows(
  lapply(c(2, 4, 8, 16), function(n_blocks) {
    dplyr::bind_rows(
      sample_prior_lag1(n_blocks, prior_on_full_corr = 0, eta = eta, seed = rbinom(1, prob = 0.3, size = 100000)),
      sample_prior_lag1(n_blocks, prior_on_full_corr = 1, eta = eta, seed = rbinom(1, prob = 0.3, size = 100000))
    )
  })
)

ggplot(prior_df, aes(x = lag1, color = Prior, fill = Prior)) +
  geom_density(linewidth = 1.1, adjust = 1.1, alpha = 0.15) +
  facet_wrap(~N_blocks, nrow = 1, labeller = label_parsed) +
  scale_color_manual(values = c(blog_colors$teal, blog_colors$purple)) +
  scale_fill_manual(values = c(blog_colors$teal, blog_colors$purple)) +
  theme_blog() +
  labs(
    x = expression("Lag-1 ACF of " * T[A]),
    y = "Density",
    title = expression("Prior on the Lag-1 ACF of " * T[A] * " (" * eta * " = 4)"),
    subtitle = expression("Block-level prior is invariant to " * N[blocks] * "; full-matrix prior shrinks toward zero as " * N[blocks] * " grows")
  ) +
  theme(legend.position = "bottom")
```

The pattern is clear: the block-level prior (teal) stays put regardless of how many blocks we add. The full-matrix prior (purple) collapses toward zero as $N_{\text{blocks}}$ grows. We see that the LKJ prior increasingly regularizes against strong temporal autocorrelation, not because the data say so, but because the prior penalizes the determinant of a larger and larger matrix.

### 5.3 Data Generation

I simulate a deliberately prior-sensitive regime with 8 blocks of size 5 and only 10 observations. The block-specific Toeplitz component is weak, while the shared component is stronger and more structured, with $s = 0.30$ placing most of the within-block signal on $T_B$. This is exactly the setting where a practitioner would want to shrink $T_A$ harder than $T_B$.

This kind of sample size is not contrived. If the observational unit is a person, classroom, clinic, or experimental batch, it is easy to end up with single-digit or low-teens sample sizes while still recording multiple repeated measurements within each unit. The structure in the covariance is then the main source of statistical leverage: without it, estimation is driven too heavily by the geometry of a large generic correlation prior rather than by defensible scientific assumptions about how the measurements are related.

```{r}
#| label: simulate

set.seed(42)

M_blocksize <- 5
N_blocks    <- 8
D           <- M_blocksize - 1
M_total     <- N_blocks * M_blocksize
N_obs       <- 10

true_s         <- 0.30
true_psi_A_raw <- c(-0.3, 0.1, 0.0, 0.0)   # weak block-specific component
true_psi_B_raw <- c(0.4, -1.1, 0.5, 0.0)   # strong shared temporal component

R_true <- build_full_R(true_psi_A_raw, true_psi_B_raw, true_s,
                        M_blocksize, N_blocks)

# Verify positive definiteness
stopifnot(all(eigen(R_true, only.values = TRUE)$values > 0))

y <- rmvnorm(N_obs, sigma = R_true)

cat("True psi_A (constrained):", round(squash_pacf(true_psi_A_raw), 3), "\n")
cat("True psi_B (constrained):", round(squash_pacf(true_psi_B_raw), 3), "\n")
cat("True lag-1 ACF(T_A):", round(pacf_to_acf(true_psi_A_raw)[2], 3), "\n")
cat("True lag-1 ACF(T_B):", round(pacf_to_acf(true_psi_B_raw)[2], 3), "\n")
cat("True within-block lag-1 correlation:",
    round(true_s * pacf_to_acf(true_psi_A_raw)[2] +
            (1 - true_s) * pacf_to_acf(true_psi_B_raw)[2], 3), "\n")
cat("True s:", true_s, "\n")
cat("Dimensions: N_obs =", N_obs, ", M =", M_blocksize,
    ", N_blocks =", N_blocks, ", P =", M_total, "\n")
```

### 5.4 Visualizing the True Correlation Matrix

```{r}
#| label: fig-true-corr
#| fig-cap: "True compound-symmetric block-Toeplitz correlation matrix. The diagonal blocks mix a weak block-specific component with a strong shared temporal component."
#| fig-height: 7
#| fig-width: 8
#| out-width: "80%"

corr_df <- expand.grid(Row = 1:M_total, Col = 1:M_total) |>
  mutate(value = as.vector(R_true))

ggplot(corr_df, aes(x = Col, y = Row, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = blog_colors$blue, mid = "white", high = blog_colors$red,
    midpoint = 0, limits = c(-1, 1), name = "Correlation"
  ) +
  scale_y_reverse() +
  coord_equal() +
  theme_blog() +
  labs(x = "Column", y = "Row",
       title = "True Correlation Matrix",
       subtitle = paste0(N_blocks, " blocks of size ", M_blocksize, ", s = ", true_s))
```

### 5.5 Head-to-Head: Domain-Informed vs Full-Matrix Priors

The key advantage of block-level priors is **independent control**. A practitioner who believes the block-specific component should be tightly regularized but the shared component should remain more flexible can set $\eta_A = 4$ and $\eta_B = 2$. The full-matrix LKJ prior has a single $\eta$ that applies everywhere.

I fit the same data with both approaches, but this is **not** a calibration-matched comparison in the sense of equal induced priors on $T_A$ and $T_B$. The block-level prior is applied directly to the component Toeplitz matrices, while the full-matrix LKJ prior is applied to the assembled $P \times P$ correlation matrix. That means the induced shrinkage on $T_A$, $T_B$, and $s$ is different by construction and depends on the full block layout. The point of the comparison is exactly that the block-level formulation can target the two components separately, whereas the full-matrix LKJ cannot.

```{r}
#| label: fit-both
#| message: false

# Fit 1: Block-level LKJ priors -- domain-informed
#   eta_A = 4 (shrink the block-specific component)
#   eta_B = 2 (leave the shared component more flexible)
fit_block <- model$sample(
  data = list(
    M_blocksize = M_blocksize, N_blocks = N_blocks,
    eta_lkj_a = 4, eta_lkj_b = 2,
    N = N_obs, K = M_total, y = y, mu = rep(0, M_total),
    prior_on_full_corr = 0
  ),
  iter_warmup = 500, iter_sampling = 500,
  parallel_chains = 4, seed = 123, refresh = 0
)

# Fit 2: Single-eta full-matrix LKJ prior on the assembled correlation matrix
fit_full <- model$sample(
  data = list(
    M_blocksize = M_blocksize, N_blocks = N_blocks,
    eta_lkj_a = 4, eta_lkj_b = 4,
    N = N_obs, K = M_total, y = y, mu = rep(0, M_total),
    prior_on_full_corr = 1
  ),
  iter_warmup = 500, iter_sampling = 500,
  parallel_chains = 4, seed = 123, refresh = 0
)
```

```{r}
#| label: diagnostics
fit_block$diagnostic_summary()
fit_full$diagnostic_summary()
```

### 5.6 Recovery of Interpretable Lag-1 Correlations and the Mixing Weight

```{r}
#| label: fig-recovery
#| fig-cap: "Posterior 90% intervals for interpretable lag-1 correlation summaries and the mixing weight under both prior approaches. Red crosses mark the true values."
#| fig-height: 8
#| fig-width: 12
#| out-width: "100%"

extract_lag1_metrics <- function(fit, prior_label) {
  draws_A <- fit$draws("psi_A_raw", format = "matrix")
  draws_B <- fit$draws("psi_B_raw", format = "matrix")
  draws_s <- as.numeric(fit$draws("s", format = "matrix")[, 1])

  lag1_A <- apply(draws_A, 1, function(x) pacf_to_acf(x)[2])
  lag1_B <- apply(draws_B, 1, function(x) pacf_to_acf(x)[2])
  lag1_within <- draws_s * lag1_A + (1 - draws_s) * lag1_B

  data.frame(
    Prior = prior_label,
    `Lag-1 ACF of T_A` = lag1_A,
    `Lag-1 ACF of T_B` = lag1_B,
    `Within-block lag-1 correlation` = lag1_within,
    `Mixing weight s` = draws_s,
    check.names = FALSE
  ) |>
    pivot_longer(-Prior, names_to = "Quantity", values_to = "value")
}

lag1_metrics <- bind_rows(
  extract_lag1_metrics(fit_block, "Block-Level LKJ"),
  extract_lag1_metrics(fit_full, "Full-Matrix LKJ")
)

quantity_levels <- rev(c(
  "Mixing weight s",
  "Lag-1 ACF of T_A",
  "Lag-1 ACF of T_B",
  "Within-block lag-1 correlation"
))

lag1_summary <- lag1_metrics |>
  group_by(Prior, Quantity) |>
  summarize(
    q5 = quantile(value, 0.05),
    median = median(value),
    q95 = quantile(value, 0.95),
    .groups = "drop"
  ) |>
  mutate(Quantity = factor(Quantity, levels = quantity_levels))

true_lag1_df <- data.frame(
  Quantity = factor(
    c("Lag-1 ACF of T_A", "Lag-1 ACF of T_B", "Within-block lag-1 correlation", "Mixing weight s"),
    levels = quantity_levels
  ),
  true_value = c(
    pacf_to_acf(true_psi_A_raw)[2],
    pacf_to_acf(true_psi_B_raw)[2],
    true_s * pacf_to_acf(true_psi_A_raw)[2] +
      (1 - true_s) * pacf_to_acf(true_psi_B_raw)[2],
    true_s
  )
)

ggplot(lag1_summary, aes(x = median, y = Quantity, color = Prior)) +
  geom_linerange(aes(xmin = q5, xmax = q95), linewidth = 2, alpha = 0.85) +
  geom_point(size = 2.5) +
  geom_point(data = true_lag1_df, aes(x = true_value, y = Quantity),
             inherit.aes = FALSE,
             color = blog_colors$red, shape = 4, size = 3, stroke = 1.5) +
  facet_wrap(~Prior) +
  scale_color_manual(values = c(blog_colors$teal, blog_colors$purple)) +
  theme_blog() +
  labs(
    x = "Value",
    y = NULL,
    title = "Recovery of Interpretable Lag-1 Correlation Summaries and s",
    subtitle = expression("Block-level: LKJ(" * eta[A] * " = 4, " * eta[B] * " = 2) versus full-matrix LKJ(" * eta * " = 4)")
  ) +
  theme(legend.position = "none")
```

### 5.7 Correlation Matrix Recovery

```{r}
#| label: fig-corr-comparison
#| fig-cap: "Recovered correlation matrices under both prior approaches."
#| fig-height: 7
#| fig-width: 14
#| out-width: "100%"

R_block <- matrix(fit_block$summary("R")$mean, M_total, M_total)
R_full  <- matrix(fit_full$summary("R")$mean, M_total, M_total)

make_corr_df <- function(R, label) {
  expand.grid(Row = 1:M_total, Col = 1:M_total) |>
    mutate(value = as.vector(R), Matrix = label)
}

lims <- c(-1, 1)

p_true <- ggplot(make_corr_df(R_true, "True"), aes(x = Col, y = Row, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = blog_colors$blue, mid = "white", high = blog_colors$red,
                       midpoint = 0, limits = lims, name = "r") +
  scale_y_reverse() + coord_equal() + theme_blog() +
  labs(title = "True", x = NULL, y = NULL)

p_block_corr <- ggplot(make_corr_df(R_block, "Block-Level"), aes(x = Col, y = Row, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = blog_colors$blue, mid = "white", high = blog_colors$red,
                       midpoint = 0, limits = lims, name = "r") +
  scale_y_reverse() + coord_equal() + theme_blog() +
  labs(title = expression("Block-Level: " * eta[A] * "=4, " * eta[B] * "=2"), x = NULL, y = NULL)

p_full_corr <- ggplot(make_corr_df(R_full, "Full-Matrix"), aes(x = Col, y = Row, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = blog_colors$blue, mid = "white", high = blog_colors$red,
                       midpoint = 0, limits = lims, name = "r") +
  scale_y_reverse() + coord_equal() + theme_blog() +
  labs(title = expression("Full-Matrix: " * eta * "=4"), x = NULL, y = NULL)

p_true + p_block_corr + p_full_corr +
  plot_layout(nrow = 1) +
  plot_annotation(
    title = "Correlation Matrix Recovery",
    theme = theme_blog()
  )
```

```{r}
#| label: fig-scatter-comparison
#| fig-cap: "Element-wise recovery of the lower triangle under both prior approaches."
#| fig-height: 5
#| fig-width: 12
#| out-width: "100%"

scatter_block <- data.frame(
  True = R_true[lower.tri(R_true)],
  Estimated = R_block[lower.tri(R_block)],
  Prior = "Block-Level LKJ"
)
scatter_full <- data.frame(
  True = R_true[lower.tri(R_true)],
  Estimated = R_full[lower.tri(R_full)],
  Prior = "Full-Matrix LKJ"
)
scatter_df <- rbind(scatter_block, scatter_full)

ggplot(scatter_df, aes(x = True, y = Estimated)) +
  geom_abline(slope = 1, intercept = 0, color = blog_colors$red,
              linewidth = 0.8) +
  geom_point(aes(color = Prior), alpha = 0.5, size = 1.2) +
  facet_wrap(~Prior) +
  scale_color_manual(values = c(blog_colors$teal, blog_colors$purple)) +
  coord_equal(xlim = c(-1, 1), ylim = c(-1, 1)) +
  theme_blog() +
  labs(x = "True Correlation", y = "Estimated Correlation",
       title = "Element-wise Correlation Recovery") +
  theme(legend.position = "none")
```

```{r}
#| label: rmse-table

block_id <- rep(1:N_blocks, each = M_blocksize)
within_mask <- outer(block_id, block_id, `==`) & lower.tri(R_true)
between_mask <- outer(block_id, block_id, `!=`) & lower.tri(R_true)

rmse_within_block <- sqrt(mean((R_true[within_mask] - R_block[within_mask])^2))
rmse_within_full  <- sqrt(mean((R_true[within_mask] - R_full[within_mask])^2))
rmse_between_block <- sqrt(mean((R_true[between_mask] - R_block[between_mask])^2))
rmse_between_full  <- sqrt(mean((R_true[between_mask] - R_full[between_mask])^2))
rmse_overall_block <- sqrt(mean((R_true[lower.tri(R_true)] - R_block[lower.tri(R_block)])^2))
rmse_overall_full  <- sqrt(mean((R_true[lower.tri(R_true)] - R_full[lower.tri(R_full)])^2))

rmse_df <- data.frame(
  Component = c("Within-block entries", "Between-block entries", "Overall"),
  `Block-Level LKJ` = c(rmse_within_block, rmse_between_block, rmse_overall_block),
  `Full-Matrix LKJ` = c(rmse_within_full, rmse_between_full, rmse_overall_full)
)

knitr::kable(rmse_df, digits = 4)
```

The RMSE split shows exactly why block-level priors matter here: they let us regularize the weak block-specific component without over-shrinking the strong shared component. In this regime the block-level prior wins on the within-block entries, the between-block entries, and the overall matrix RMSE.

### 5.8 Posterior on the Within-Block Lag-1 Correlation and s

The most practically relevant quantity is the lag-1 correlation inside a diagonal block: $$
\rho_{\text{within}, 1} = s \rho_{A, 1} + (1 - s) \rho_{B, 1}.
$$ This is what determines how strongly adjacent time points within the same subject are correlated. The other key scalar is $s$, which controls how much of that within-block dependence is attributed to the block-specific component versus the shared component. With $\eta_A = 4$ the block-level prior shrinks the weak block-specific component, while $\eta_B = 2$ leaves the shared component flexible enough to match the data. The full-matrix LKJ(4) prior cannot separate these roles, so it ends up shrinking both the lag-1 summary and the implied decomposition differently.

```{r}
#| label: fig-lag1-s-posterior
#| fig-cap: "Posterior distributions of the actual within-block lag-1 correlation and the mixing weight s under both priors."
#| fig-height: 6
#| fig-width: 10
#| out-width: "90%"

draws_block_A <- fit_block$draws("psi_A_raw", format = "matrix")
draws_block_B <- fit_block$draws("psi_B_raw", format = "matrix")
draws_block_s <- as.numeric(fit_block$draws("s", format = "matrix")[, 1])

draws_full_A <- fit_full$draws("psi_A_raw", format = "matrix")
draws_full_B <- fit_full$draws("psi_B_raw", format = "matrix")
draws_full_s <- as.numeric(fit_full$draws("s", format = "matrix")[, 1])

lag1_A_block <- apply(draws_block_A, 1, function(x) pacf_to_acf(x)[2])
lag1_B_block <- apply(draws_block_B, 1, function(x) pacf_to_acf(x)[2])
lag1_A_full  <- apply(draws_full_A, 1, function(x) pacf_to_acf(x)[2])
lag1_B_full  <- apply(draws_full_B, 1, function(x) pacf_to_acf(x)[2])

lag1_block  <- draws_block_s * lag1_A_block + (1 - draws_block_s) * lag1_B_block
lag1_full   <- draws_full_s * lag1_A_full  + (1 - draws_full_s) * lag1_B_full

true_lag1 <- true_s * pacf_to_acf(true_psi_A_raw)[2] +
  (1 - true_s) * pacf_to_acf(true_psi_B_raw)[2]

lag1_df <- data.frame(
  value = c(lag1_block, lag1_full, draws_block_s, draws_full_s),
  Prior = c(
    rep(c("Block-Level LKJ", "Full-Matrix LKJ"), each = length(lag1_block)),
    rep(c("Block-Level LKJ", "Full-Matrix LKJ"), each = length(draws_block_s))
  ),
  Quantity = c(
    rep("Within-block lag-1 correlation", 2 * length(lag1_block)),
    rep("Mixing weight s", 2 * length(draws_block_s))
  )
)

true_posterior_df <- data.frame(
  Quantity = c("Within-block lag-1 correlation", "Mixing weight s"),
  true_value = c(true_lag1, true_s)
)

ggplot(lag1_df, aes(x = value, fill = Prior)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 alpha = 0.6, position = "identity") +
  geom_vline(data = true_posterior_df, aes(xintercept = true_value),
             color = blog_colors$red, linewidth = 1,
             inherit.aes = FALSE) +
  scale_fill_manual(values = c(blog_colors$teal, blog_colors$purple)) +
  facet_grid(Quantity ~ Prior, scales = "free_x") +
  theme_blog() +
  labs(
    x = "Value",
    y = "Density",
    title = "Posterior on the Within-Block Lag-1 Correlation and s",
    subtitle = "Red line = true value"
  ) +
  theme(legend.position = "none")
```

## 6. When to Use Which Prior

| Scenario | Recommended Prior |
|:-----------------------------------|:-----------------------------------|
| Domain knowledge that within- and between-block correlation differ | Block-level: set $\eta_A \neq \eta_B$ |
| Many blocks ($N$ large) | Block-level: avoids $N$-dependent shrinkage on $T_A$ |
| Want a single "regularize toward identity" prior and $P$ is small | Full-matrix LKJ is fine |
| No block structure at all | Standard LKJ (this model doesn't apply) |

There is no free lunch, the full-matrix LKJ prior provides a single style of regularization, which can still be useful when there is little structural knowledge. The block-level prior's advantage is not about always having lower error, it's about **encoding domain knowledge that the full-matrix prior cannot represent**. When a practitioner believes the block-specific and shared components should be regularized differently, the block-level prior translates that directly into $\eta_A \neq \eta_B$. The full-matrix prior has no mechanism for this asymmetry.

## 7. Discussion

The key insight is structural: **when a correlation matrix decomposes into interpretable components, the prior should too**. The PACF parameterization enables this by giving each Toeplitz block an independent parameter vector with a clean Jacobian. The compound-symmetric block-Toeplitz structure then provides the scaffold that turns two small priors into a valid prior on the full $P \times P$ matrix.

The practical value is prior elicitation. In most panel or longitudinal settings, a researcher has different beliefs about temporal dependence within subjects versus cross-sectional dependence between subjects. Block-level priors make these beliefs expressible; the full-matrix LKJ prior forces a single $\eta$ that governs everything.

An alternative route is to place a prior directly on the Toeplitz row through a spectral parameterization, such as a finite mixture of AR(1) components, and pair it with Trench-style Toeplitz linear algebra [@trench1985inverse; @trench1985eigen]. Compared with that approach, the model in this post makes a different tradeoff. The spectral formulation is natural when the main object of interest is a single Toeplitz covariance and the prior should act directly on its spectrum. Here the goal is different: we want a block-specific Toeplitz component and a shared Toeplitz component, each with its own prior strength, combined through one scalar mixing parameter. The PACF parameterization makes each Toeplitz block valid by construction, while the compound-symmetric decomposition makes the full block matrix valid by construction. That is why this formulation is a better fit for prior elicitation on "within-block" versus "shared-across-block" correlation, even though a Trench-based implementation would be a strong alternative for pure Toeplitz models.

This principle generalizes beyond the specific structure shown here. Any time a large correlation matrix is built from smaller pieces -- Kronecker products, block-diagonal structures, factor models -- it's worth asking whether priors on the pieces give better behavior than a single prior on the assembly.

The Stan implementation exposes both options via a single flag (`prior_on_full_corr`), making it easy to compare in practice. The efficient innovations-based likelihood handles the computation regardless of which prior is used.

## Appendix: Stan Code

``` {.stan filename="compound-symmetric-block-toeplitz.stan"}
{{< include compound-symmetric-block-toeplitz.stan >}}
```